GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2024-12-03 writeUps/paper.tex
Updates from Overleaf

\documentclass{statsoc} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{amsmath,amssymb}
\usepackage{bm}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}
%\usepackage{multibib}

@@

%\newcites{supp}{Supplementary References}

%%% TO DO
%% Add detail to application section
%% fill in missing citations
%% Appendix
%%   clean up/fix notation for math
%%   add in full simulation results
%%   small interaction simulation
%%   look for what else I promised and either put it in or delete promise
%% proofread

\title[\textsc{geepers}]{\textsc{geepers}: Principal Stratification using Principal Scores and Stacked Estimating Equations}

\author[Sales et al.]{Adam C Sales}

\author{Kirk P Vanacore}
\author{Erin R Ottmar}

\address{Worcester Polytechnic Institute,
Worcester, MA,
USA.}
\email{asales@wpi.edu}

\begin{document}
\maketitle

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and to check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We illustrate the new technique in an analysis of student log data from a recent educational technology field experiment.  
\end{abstract}
\keywords{Causal Inference; Principal Stratification; Educational Technology}

@@

Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
%A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline %(such as OLS with interactions) 
will often produce inconsistent estimates.

Principal stratification \citep{frangakis} is a broad framework for defining subgroup effects for post-treatment subgroups. It relies on the notion of \emph{potential} subgroups called principal strata: groups of subjects defined based on the values a post-treatment variable would potentially take if the subject were assigned to one treatment condition or another.
Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignments, it is in general unobserved.
Nonetheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems, including non-compliance in randomized experiments \citep{air}, the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, and sample attrition \citep{zhangRubin, ding2011}.

Because principal strata membership is generally unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbens1997bayesian}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased and misleading \citep{griffin2008application,feller2016principal}.
Another route to identification and estimation of principal effects leads through auxiliary data, including baseline covariates or secondary outcomes \citep{mattei2013exploiting}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute form an unbiased estimate of principal effects.

On the other hand, \citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on a general estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
It turns out that, under certain general conditions, principal effects can be estimated with a simple, two-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores and estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``\textsc{geepers}'': generalized estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on correct specification of regression functions, but unlike common fully-parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability, and admits to sandwich-style standard error estimates, which we provide in the appendix.

This paper focuses on a special case of principal stratification---one-way, binary non-compliance---that is, there are only two principal strata, and principal stratum membership is observable in one treatment arm, but not in the other.
This scenario would occur if, say, subjects in the control arm of a study had no access to the treatment, and not all subjects assigned to the treatment arm actually received the full treatment.
While such a setup may represent the simplest and easiest principal stratification problem, we believe it is a common scenario.
Moreover, we believe that the methods developed here can extend to more complex scenarios, which we leave to future work.

The following section gives a formal introduction to principal stratification and briefly describes two common estimation techniques, parametric mixture modeling and principal score weighting.
Next, Section \ref{sec:geepers} develops \textsc{geepers} estimation, and gives conditions for its consistency.
Section \ref{sec:simulation} describes a simulation study examining the operating characteristics of \textsc{geepers} estimates and comparing them to estimates from parametric mixture modeling and principal score weighting.
Under ideal circumstances, estimates from parametric mixture models and weighting estimators are somewhat more efficient than \textsc{geepers}, but when assumptions are violated \textsc{geepers} easily out-performs the other two methods.
Section \ref{sec:fh2t} illustrates \textsc{geepers} in analyzing data from a recent large-scale field trial of educational software, and Section \ref{sec:discussion} concludes.

\section{Background}
In a randomized experiment with two conditions, let $Z_i=0$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken subsequent to treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff1=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is due to the fact that strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

We will focus on estimation in completely randomized experiments, that is, we will assume
\begin{ass}[Randomization]\label{ass:rand}
\begin{equation*}%\label{eq:randomization}
  \yci,\yti,\sti, \bm{x}_i \independent Z_i
\end{equation*}
\end{ass}
where $\bm{x}_i$ is a vector of pre-treatment covariates.

@@

This approach has a number of setbacks.
First, the conditional normality assumption is untestable and limiting.
Unfortunately, even when it holds, mixture model estimates can be unreliable and, in some cases, severely biased \citep{griffin2008application,feller2017principal}.

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=1$.

@@

In order for principal scores to be useful, they need to vary with $\bm{x}$. Otherwise, $\bm{x}$ is uninformative about $\st$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
%In some estimators we will discuss below, two distinct values suffice. 
For a broader discussion, see \citet{ding2011,jiangDing2021}.

@@

\section{\textsc{geepers}}\label{sec:geepers}

The approach we introduce here incorporates principal scores into an M-Estimator for the mixture model \eqref{eq:mixtureModel}.
Like principal score weighting, our M-estimator does not require distributional assumptions, but instead requires a conditional independence assumption.
We will begin by describing a stronger-than-necessary conditional independence assumption in order to build intuition; in \S~\ref{sec:regression} we will present a considerably weaker alternative.

@@

We begin by introducing an assumption that we term ``Covariate Ignorability,'' or CI. It is a slightly weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Ignorability]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how, we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS. % assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

@@

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions \citep[e.g.][p. 327]{boosStefanskiBook}, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  %\begin{split}
    \heff0_{CI}\equiv \hat{\beta}_2\mbox{ and }
    \heff1_{CI}\equiv \hat{\beta}_3+\hat{\beta}_2
%  \end{split}
\end{equation}
are M-estimators, with $\heff0_{CI}\rightarrow_p\eff0$ and $\heff1_{CI}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

\subsection{Relaxing Covariate Ignorability with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).

The assumption we'll state here is stronger than necessary, but leads to an attractively simple method; a weaker version of the assumption and corresponding method, along with the proof, can be found in an online appendix.

\begin{ass}[Residualized Covariate Ignorability]\label{ass:rci}
There exists $\bxy$, a (known) transformation of covariates $\bx$, and an (unknown) vector of coefficients $\bm{\gamma}$ such that
\begin{equation}\label{eq:ass:rci}
\begin{split}
\EE[\yc-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yc-\bm{\gamma}'\bxy|\st]\\
\EE[\yt-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yt-\bm{\gamma}'\bxy|\st]
\end{split}
\end{equation}
\end{ass}
That is, CI may not hold for potential outcomes themselves, but an analogous assumption holds for \emph{residualized} potential outcomes---i.e. $\yc$ and $\yt$ after subtracting out a linear function of (possibly transformed) covariates.
By forcing $\bm{\gamma}$ to be the same for both treatment groups, and (implicitly) for both principal strata, Assumption \ref{ass:rci} assumes that there are no interactions between columns of $\bxy$ and either condition or principal stratum.
This is akin to the ``Additivity of Treatment Assignment Effect'' assumption in \citet{jo2002}.
Proposition \ref{prop:interactions} in the online appendix relaxes these requirements, but some preliminary simulations suggest that standard errors from estimates allowing for these interactions will be prohibitively large. 
Simulation results in the following section address some cases in which interactions are present in the data generating model but not in the data analysis.

@@

The distribution of $\epsilon_i$, $\beta_1$, and coefficients $\gamma_1$, $\gamma_2$, and $\gamma_3$ were manipulated factors in the study.

@@

Using each simulated dataset, we estimated $\eff0$ and $\eff1$ with three methods: a mixture model \eqref{eq:mixtureModelPS} including principal scores, \textsc{geepers}, and third, the principal score weighting estimator \eqref{eq:psw}.

@@

\clearpage
\setcounter{page}{1}

>>>>

\documentclass[]{interact}

\RequirePackage{amsthm,amsmath,amsfonts,amssymb}
%\RequirePackage[authoryear]{natbib}
\usepackage{longtable}
\usepackage[notablist,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage[caption=false]{subfig}% Support for small, `sub' figures and tables
\usepackage[doublespacing]{setspace}% To produce a `double spaced' document if required
%\setlength\parindent{24pt}% To increase paragraph indentation when line spacing is doubled

\usepackage[longnamesfirst,sort]{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{,}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% To set the list of references in 10 point font using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by asthma.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{prop}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)
\usepackage{bm}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{mathtools}
%\usepackage{fullpage}
\usepackage{makecell}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{numdef}
%\usepackage{multibib}
\usepackage{rotating}

%\startlocaldefs

@@

%\endlocaldefs
%\newcites{supp}{Supplementary References}
\begin{document}

%\begin{frontmatter}

\title{GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations}
%\runtitle{\textsc{geepers}}
%\maketitle

%\begin{aug}
\author{
\name{Adam C. Sales,\thanks{CONTACT Adam C. Sales. Email: asales@wpi.edu} Kirk P. Vanacore, and Erin R. Ottmar}
\affil{Worcester Polytechnic Institute, Worcester, MA, USA}
}

\maketitle

\begin{abstract}
Principal stratification is a framework for making sense of causal effects conditioned on variables that themselves may have been affected by treatment. For instance, one component of an educational computer application is the availability of “bottom-out” hints that provide the answer. In evaluating a recent experimental evaluation against alternative programs without bottom-out hints, researchers may be interested in estimating separate average treatment effects for students who, if given the opportunity, would request bottom-out hints frequently, and for students who would not. Most principal stratification estimators rely on strong structural or modeling assumptions, and many require advanced statistical training to fit and check. In this paper, we introduce a new M-estimation principal effect estimator for one-way noncompliance based on a binary indicator. Estimates may be computed using conventional regressions (though the standard errors require a specialized sandwich formula) and do not rely on distributional assumptions. We present a simulation study that shows that the novel method is more robust than popular alternatives and illustrate the method in an analysis of data on bottom-out hint requests. 
\end{abstract}
\begin{keywords}
Causal Inference; Principal Stratification; Educational Technology
\end{keywords}
%\end{frontmatter}

\clearpage

@@

%Randomized experiments easily admit subgroup analysis---estimation of treatment effects for a subset of the study population---when those subgroups are defined at baseline, and not themselves a function of treatment assignment.
%A simple, popular approach to estimating subgroup effects is to use ordinary least squares (OLS) regression, regressing the outcome on a treatment indicator interacted with an indicator for subgroup membership.
%However, when subgroup membership is a function of a post-treatment variable, then it may itself be affected by the treatment.
%Contrasts between treatment and control conditions within those subgroups may not have a causal interpretation at all---and when they do, the same techniques that work for subgroups defined at baseline %(such as OLS with interactions) 
%will often produce inconsistent estimates.

%%% open with hook for ed researchers

When students get stuck on math problems during homework or classwork, it may be helpful to offer them hints or other ``just-in-time'' support \cite[e.g][]{jit}, possibly helping them overcome specific misconceptions or other sources of confusion and allowing them to move on in the assignment. 
The availability of problem-specific support for struggling students is one potential advantage of educational technology over traditional pencil-and-paper practice. 
On the other hand, the ready availability of hints may discourage students from working through difficult problems on their own, and so may inhibit learning.

The literature on hints in online learning is mixed \citep[see, e.g.,][]{aleven2016help,goldin2012learner,sales2021student} in particular when it comes to requesting a ``bottom-out'' hint. In some online mathematics programs, sequences of hints, which a student can request one at a time, end with a bottom-out hint providing the answer to the problem. On the one hand, hints that include the correct answer are essentially worked examples, which can be beneficial to learning \citep[e.g.]{sweller1985use}. On the other hand, they allow students to ``game the system,'' by simply requesting enough hints to see the right answer, and use the software without ever working to solve any problems \citep[e.g.][]{guo2008trying}.

Data from the randomized controlled trial (RCT) reported in \citet{impactPaper} may provide evidence for one or another side of this debate. 
That RCT contrasted four different applications for computer-based math practice, one of which included bottom-out hints. 
If that condition benefited students who frequently requested bottom-out hints (``bottom-outers'') more than those who didn't, that may suggest that bottom-out hints can be helpful. 
If students who avoided bottom-out hints benefited more, it may be because bottom-out hints were harmful.
(There are certainly other factors at play as well, so this analysis would not be conclusive.)

Estimating an effect for bottom-outers is a tricky proposition. 
Causal inference in statistics typically depends on comparisons of outcomes between similar groups of subjects. However, there is no readily identifiable comparison group for bottom-outers, who are only represented in one of the four treatment conditions, and who presumably struggled more frequently, on average, than their non-bottom-outer peers. 
We will formulate a solution using Principal stratification \citep{frangakis}, a  framework for defining subgroup effects for subgroups that may themselves be affected by treatment assignment. It relies on the notion of \emph{potential} subgroups called principal strata, grouping subjects based on the values a post-treatment variable would potentially take if the subject were assigned to one treatment condition or another.
For example, the appropriate comparison group for bottom-outers is the set of students, assigned to the other treatment arms, who \emph{would have} requested bottom-out hints if only given that option.

Since principal strata membership is based on subjects' responses to both factual and counter-factual treatment assignments, it is unobserved---we don't observe which students in the conditions without bottom-out hints would have bottomed out if given the opportunity.
Nonetheless, researchers have devised estimators of principal effects, as average effects within principal strata are called, to address a broad range of applied statistical problems.
For instance, \citet{air} introduced the framework (or an antecedent thereof) for the case of non-compliance, in which treatment assignment and treatment receipt don't match. They applied it to estimate the effects of military service in the American war in Vietnam using the draft lottery, even though some draftees didn't serve and some people served without being drafted, by focusing on the principal stratum of Americans who would serve only if drafted.
\citet{fellerEtAl2016} estimated differential effects of random assignment to the Head Start early childhood care program for children who, if assigned to control, would receive home-based care, and for those who would attend a different early childhood care program. 
Other methodologists have developed estimators for the evaluation of surrogate outcomes \citep{li2010bayesian}, the investigation of causal mechanisms \citep{lidsayPage}, sample attrition \citep{zhangRubin, ding2011}, and other problems.

Because principal strata membership is unobserved, principal effects are not always identified, and when they are, estimation can still be difficult.
Typically, principal effects are identified and estimated based on strong structural assumptions \citep[e.g.][]{air} and/or parametric modeling \citep[e.g.][]{imbens1997bayesian}.
Parametric models for principal effects are often hard to specify and hard or impossible to test.
Even when they are well-specified they can yield estimates that are severely biased  \citep{griffin2008application,feller2016principal}.
The most common type of parametric model, the normal mixture model, is especially dubious for studying bottom-outers in \citet{impactPaper}'s RCT, whose outcome, the total score on a 10-item posttest, is not normally distributed.

Another route to identification and estimation of principal effects leads through auxiliary data, including baseline covariates or secondary outcomes \citep{mattei2013exploiting}.
Of particular importance is the ``principal score,'' \citep{jo,dingLu,feller2017principal}, or the probability that a subject belongs to a particular principal stratum, as a function of pre-treatment covariates.
Under a strong, untestable assumption called ``principal ignorability,''  principal scores can be transformed into weights to compute an unbiased estimate of principal effects.

%%% put in paragraph here or above with PS in ed research (lindsay page, feller, my papers)

\citet{jo2002}, \citet{ding2011}, \citet{jiangDing2021}, and others have shown that, under certain circumstances, researchers can use baseline covariates to identify and estimate principal effects without relying on principal ignorability.
This paper expands on those identification results by presenting a new set of principal effect estimators, and associated standard error estimates, based on general estimating equations (or M-estimation or method-of-moments) \citep{stefanskiBoos}.
Under certain general conditions, principal effects can be estimated with a simple, three-step procedure: first, estimate principal scores with logistic regression or another M-estimator; then, impute unknown principal subgroup membership indicators with principal scores, and finally estimate principal effects using ordinary least squares (OLS) regression with interactions. %, just as in the pre-treatment subgroup case.
We will refer to the estimators presented here as ``\textsc{geepers}'': general estimating equations for principal effects using regressions.
\textsc{geepers} estimates rely on the correct specification of regression functions, but unlike common fully parametric methods, they make no assumptions about the shape of the distribution of regression errors.
Unlike principal score weighting, \textsc{geepers} does not assume principal ignorability and admits to sandwich-style standard error estimates, which we provide in the appendix.

This work advances the literature in three ways: first, it makes an explicit connection between principal score methods, M-estimation, and semi-parametric estimation of principal effects using auxiliary data. Second, it provides a general estimator that relies on two of the most familiar tools in the applied education researcher's toolbox: linear and logistic regression. The \textsc{geepers} method can, therefore, make principal stratification methods accessible for a much broader range of applied education researchers. Lastly, this work includes an extensive simulation study that compares \textsc{geepers} to parametric mixture modeling and principal score weighting under a range of favorable and challenging scenarios. The simulation demonstrates the strengths and weaknesses of each method and the tradeoffs between them.

This paper focuses on a special case of principal stratification---one-way, binary non-compliance---that is, there are only two principal strata and principal stratum membership is observable in one treatment arm, but not in the other.
This scenario would occur if, say, subjects in the control arm of a study had no access to the treatment, and not all subjects assigned to the treatment arm received the full treatment.
While such a setup may represent the simplest and easiest principal stratification problem, we believe it is a common scenario.
For example, \textsc{geepers} could be used to estimate effects from an RCT for participants categorized as low- or high-implementers. It could also be used to assess more granular questions about effect heterogeneity, such as whether an intervention's impact is contingent upon specific behaviors displayed by either practitioners or students.

The following section gives a formal introduction to principal stratification and briefly describes two common estimation techniques, parametric mixture modeling and principal score weighting.
Next, Section \ref{sec:geepers} develops \textsc{geepers} estimation and gives conditions for its consistency.
Section \ref{sec:simulation} describes the simulation study. % examining the operating characteristics of \textsc{geepers} estimates and comparing them to estimates from parametric mixture modeling and principal score weighting.
%Under ideal circumstances, estimates from parametric mixture models and weighting estimators are somewhat more efficient than \textsc{geepers}, but when assumptions are violated \textsc{geepers} easily out-performs the other two methods.
Section \ref{sec:fh2t} illustrates \textsc{geepers} by returning to the question of bottom-out hints: does the effect of being randomized to a condition that includes bottom-out hints depend on how often they are requested?  Section \ref{sec:discussion} concludes.

\section{Background}
In a randomized experiment with two conditions, let $Z_i=1$ if subject $i$, $i=1,\dots,n$ is assigned to the treatment condition, and let $Z_i=0$ otherwise.
If interest is in the effect of $Z$ on an outcome $Y$, under the ``stable unit treatment value assumption'' \citep{air} of no interference between units and no hidden versions of the treatment, define potential outcomes \citep{splawa1990application} $\yti$ as the value of $Y$ that $i$ would exhibit if $Z_i=1$, and let $\yci$ be the value if $Z_i=0$. Then the observed value of $Y$ satisfies $Y_i=Z_i\yti+(1-Z_i)\yci$, and the treatment effect of $Z$ on $Y$ for subject $i$ may be defined as $\tau_i=\yti-\yci$. The average treatment effect (ATE) is $\EE[\tau]$. %

Let $S_i\in\{0,1\}$ be a measurement on subject $i$ taken after treatment assignment, so that $S$ may be affected by $Z$.
Then $S$ itself has potential values $\sti$ and $\sci$, which $i$ would exhibit if $Z_i=1$ or $Z_i=0$, respectively.
Assume ``strong monotonicity'' \citep[c.f.][]{dingLu} (also referred to as ``one-way noncompliance''):%\citealt{wing2017can}):
\begin{ass}[Strong Monotonicity]\label{ass:sm}
\begin{equation*}%\label{eq:strongMonotonicity}
  \sci=0\text{ for all }i
\end{equation*}
\end{ass}
Like potential outcomes $\yti$ and $\yci$, $\sci$ and $\sti$ are defined for all subjects, though $\sci$ is only observed when $Z_i=0$ and $\sti$ is only observed when $Z_i=1$. Under strong monotonicity, since $S$ is binary, every subject belongs to one of two ``principal strata,'' $\sti=1$ or $\sti=0$.
The average treatment effect in each principal stratum, $\eff1=\EE[\tau|\st=1]$ or $\eff0=\EE[\tau|\st=0]$, is called a principal effect. %
The challenge in estimating principal effects is because strata membership $\sti$ is only observed when $Z_i=1$, in which case $\sti=S_i$, the observed value of $S$.

@@

We will focus on estimation in completely randomized experiments, that is, we will assume
\begin{ass}[Randomization]\label{ass:rand}
\begin{equation*}%\label{eq:randomization}
  \yci,\yti,\sti, \bm{x}_i \independent Z_i
\end{equation*}
\end{ass}
where $\bm{x}_i$ is a vector of pre-treatment covariates, and $\independent$ denotes independence.

@@

This approach has several problems.
First, the conditional normality assumption is untestable and limiting.
Unfortunately, even when it holds, mixture model estimates can be unreliable and, in some cases, severely biased \citep{griffin2008application,feller2017principal}.

\subsection{Principal Scores}
The principal score \citep[e.g.][]{jo} is defined as
\begin{equation}\label{eq:pscore}
  \ppi=Pr(\sti=1|\bxsi)
\end{equation}
where $\bxsi\subseteq\bm{x}_i$, the set of covariates used to model $S$.
Note that under randomization, $Pr(\sti=1|\bxsi,Z_i)=Pr(\sti=1|\bxsi)$, so a model for principal scores can be estimated using data from the treatment group and extrapolated to the control group.
Assume a model for principal scores:
\begin{equation}\label{eq:prinScoreMod}
  \ppip{\bm{\alpha}}=f(\bxsi;\bm{\alpha})
\end{equation}
with parameter vector $\bm{\alpha}$. An analyst may estimate $\bm{\alpha}$ by fitting model \eqref{eq:prinScoreMod} using observed $\bm{S}$ and $\bxs$ for subjects with $Z=1$, and then compute estimated principal scores $\ppip{\bm{\hat{\alpha}}}$ for subjects with $Z_i=0$.

@@

For principal scores to be useful, they need to vary with $\bm{x}$. Otherwise, $\bm{x}$ is uninformative about $\st$. In particular, we will assume the following:
\begin{ass}[Variable Principal Scores]\label{ass:vps}
 $\pp$ takes at least 3 distinct values
\end{ass}
%In some estimators we will discuss below, two distinct values suffice. 
For a broader discussion, see \citet{ding2011,jiangDing2021}.

@@

\section{GEEPERS }\label{sec:geepers}

The approach we introduce here incorporates principal scores into an M-Estimator for the mixture model \eqref{eq:mixtureModel}.
Like principal score weighting, our M-estimator does not require distributional assumptions but instead requires a conditional independence assumption.
We will begin by describing a stronger-than-necessary conditional independence assumption to build intuition; in \S~\ref{sec:regression} we will present a considerably weaker alternative.

@@

We begin by introducing an assumption that we term ``Covariate Independence,'' or CI. It is a slightly weaker version of \citet{jiangDing2021}'s ``auxiliary independence'' assumption.
\begin{ass}[Covariate Independence]\label{ass:ci}
\begin{equation}\label{eq:assumption}
\EE[\yci|\bxsi,\sti]=\EE[\yci|\sti]=\muc0\text{ or }\muc1
\end{equation}
\end{ass}
i.e. $Y_C$ is mean-independent of $\bxs$ conditional on $\st$.
Under CI, covariates are not informative of the mean of $Y_C$ within principal strata.
As stated above, CI will rarely be plausible.
However, in some circumstances, researchers can identify a subset of observed covariates that satisfy CI, and use those covariates in estimation.

It turns out that under strong monotonicity, randomization, and CI, and given a set of principal scores, the two principal effects $\eff0$ and $\eff1$ can be estimated by a simple OLS regression.
To see how we'll %first
derive the estimating equations for $\muc0$, $\muc1$, $\mut0$, and $\mut1$, and show that they are equivalent to estimating equations for a model fit by OLS. % assuming that principal scores are known exactly, and then generalize to the case in which they are estimated using data from the treatment group.
The details of the calculations and the proofs are in the appendix.

@@

\begin{prop}\label{prop:reg1}
  Under Assumptions \ref{ass:sm}, \ref{ass:rand}, \ref{ass:vps}, \ref{ass:ci}, and suitible regularity conditions, if %principal scores are known exactly,
  $\ri$ is defined as in \eqref{eq:ri} and the model
\begin{equation}\label{eq:regression0}
  \EE[Y_i|Z_i,\ri]=\beta_0+\beta_1\ri+\beta_2Z_i+\beta_3Z_i\ri
\end{equation}
is fit with OLS, yielding coefficient estimates $\hat{\beta}_k$, $k=0,\dots,3$, then
\begin{equation}
  %\begin{split}
    \heff0_{CI}\equiv \hat{\beta}_2\mbox{ and }
    \heff1_{CI}\equiv \hat{\beta}_3+\hat{\beta}_2
%  \end{split}
\end{equation}
are M-estimators, with $\heff0_{CI}\rightarrow_p\eff0$ and $\heff1_{CI}\rightarrow_p\eff1$ as $n\rightarrow\infty$
\end{prop}

@@

\subsection{Relaxing Covariate Independence with Outcome Modeling}\label{sec:regression}
In most applications, identifying a set of covariates $\bxs$ satisfying CI will be difficult or impossible.
Fortunately, CI may be relaxed or obviated by regression (see \citealt[][\S 3.4]{jiangDing2021} for an analogous identification result).

The assumption we'll state here is stronger than necessary, but leads to an attractively simple method; a weaker version of the assumption and corresponding method, along with the proof, can be found in an appendix.

\begin{ass}[Residualized Covariate Independence]\label{ass:rci}
There exists $\bxy$, a (known) transformation of covariates $\bx$, and an (unknown) vector of coefficients $\bm{\gamma}$ such that
\begin{equation}\label{eq:ass:rci}
\begin{split}
\EE[\yc-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yc-\bm{\gamma}'\bxy|\st]\\
\EE[\yt-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yt-\bm{\gamma}'\bxy|\st]
\end{split}
\end{equation}
\end{ass}
That is, CI may not hold for potential outcomes themselves, but an analogous assumption holds for \emph{residualized} potential outcomes---i.e. $\yc$ and $\yt$ after subtracting out a linear function of (possibly transformed) covariates.
By forcing $\bm{\gamma}$ to be the same for both treatment groups, and (implicitly) for both principal strata, Assumption \ref{ass:rci} assumes that there are no interactions between columns of $\bxy$ and either condition or principal stratum.
This is akin to the ``Additivity of Treatment Assignment Effect'' assumption in \citet{jo2002}.
Proposition \ref{prop:interactions} in the appendix relaxes these requirements, but some preliminary simulations suggest that standard errors from estimates allowing for these interactions will be prohibitively large. 
Simulation results in the following section address some cases in which interactions are present in the data-generating model but not in the data analysis.

@@

The distribution of $\epsilon_i$, $\beta_1$, and values of coefficients $\gamma_1$, $\gamma_2$, and $\gamma_3$ were manipulated factors in the study.

@@

Using each simulated dataset, we estimated $\eff0$ and $\eff1$ with three methods: a mixture model \eqref{eq:mixtureModelPS} including principal scores, \textsc{geepers}, and the principal score weighting estimator \eqref{eq:psw}.

@@
